Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.tex

  \title{Examining spatial equity in accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[Some School]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[Another University]{Christopher D. Higgins\corref{2}}
   \ead{cd.higgins@utoronto.ca} 
    \author[Some School]{Antonio Paez\corref{2}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
M1C1A4}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  Public bicycle share programs (PBSPs) are implemented to promote cycling
  as a convenient and sustainable mode of transportation. These systems
  can also play a role in addressing transportation needs and advancing
  transportation equity if they make cycling more accessible to lower
  income or disadvantaged populations. In Ontario, the City of Hamilton
  launched a public bicycle share program in 2015 that currently has over
  900 operational bicycles and 130 docking stations. This system was
  expanded in 2018 by an equity initiative that added twelve docking
  stations with the explicit objective of increasing spatial equity in
  access. Since the cost of reaching a bicycle share station decreases the
  potential of accessing the program, the location of stations matters,
  and previous research found that Hamilton's public bicycle share program
  targeted well disadvantaged areas in the city. In this research, we
  revisit the case of Hamilton and investigate differentials in
  accessibility to bicycle share stations using a balanced floating
  catchment area (BFCA) accessibility approach. We compare accessibility
  with and without the equity stations to assess the effect of the
  initiative. We implement our analysis parting from micro zones to better
  reflect walking to a docking station, and conduct a sensitivity analysis
  at several walking time thresholds. We then reaggregate the estimated
  accessibility for further analysis using census data. The addition of
  equity stations increased the serviced population at every threshold
  examined, and although accessibility increased for the whole population,
  this increase was relatively modest especially for population in the
  bottom 20\% of median household income.
  \end{abstract}

@@

The potential of public bicycle share programs (PBSPs) to increase
cycling levels is but one of many reasons for implementing such programs
in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
and convenient form of public transportation, shared bicycles can
encourage individuals to take up cycling for short local trips or first
and last mile trips to other public transportation instead of using
personal vehicles. These systems can also play a role in addressing
transportation needs and advancing transportation equity if they make
cycling more accessible to lower income or disadvantaged populations. As
these systems become increasingly common, their introduction has been
accompanied by a flurry of research that investigates them from the
perspective of spatial and transportation equity (e.g., Smith et al.,
2015; Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et
al., 2019; Qian and Jaller, 2020; Qian et al., 2020). Indeed, although
these programs are available to the general public and ought to be
accessible to any individuals who wish to use them, research on PBSPs
located in North American and European cities indicates that there are
inequities in spatial accessibility in need of redress.

In this paper we investigate the case of the PBSP in the City of
Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known
as Social Bicycles, or SoBi) was launched in 2015 and currently has over
900 operational bicycles and 130 docking stations. An interesting
feature of this system is that it was expanded by an equity initiative
that introduced twelve docking stations in 2018 with the explicit
objective of increasing spatial accessibility. Hamilton Bike Share was
already studied by Hosford and Winters (2018) as part of a selection of
PBSPs in Canada. These authors found that most of the cities in their
research could benefit of greater efforts to expand service to areas
with lower socio-economic populations - but Hamilton fared somewhat
better than other cities in this respect. What distinguishes our
research is the use of a balanced floating catchment area (BFCA)
accessibility approach that accounts for the supply of stations and the
potential demand from the population serviced. This requires a more
disaggregated approach than the use of census geographies (q.v., Hosford
and Winters, 2018), since many trips to a bike share station likely
happen at a level lower than even the smallest census geography. For
this reason, we implement our analysis parting from micro population
zones to better reflect the friction of walking to a docking station.
Further, we conduct a sensitivity analysis at several walking time
thresholds. In terms of spatial equity, we compare accessibility with
and without the equity stations to assess the effect of the initiative,
before reaggregating the data for further analysis using median
household income information from the census. We find that the addition
of equity stations does increase the serviced population at every
threshold examined. However, although accessibility increased for the
whole population, this increase was relatively modest, especially for
population in the bottom 20\% of median household income. The analysis
also identifies areas under-serviced within the core service area of the
system.

@@

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent interest in prioritizing local proximity and modes
that are suitable for local trips like walking and cycling (Levine,
2020). As a measure of the ease of reaching potential destinations
spread spatially in a given area, accessibility is relevant to PBSPs
because it can identify current inequities in the provision of
infrastructure, as well as guide interventions that increase access for
groups that are under-serviced or address gaps in transportation
options. It also addresses some of the challenges of other performance
measures such as level of service within a transportation network by
measuring person-based indicators and exploring differences in use
between population groups (Páez et al., 2012).

Beyond the utility derived from using shared bikes to destinations of
value, an important aspect of accessibility associated with PBSPs is the
distance an individual must travel to reach a bicycle share station
(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
distance needed to reach a bicycle share station decreases the potential
of accessing the program and ultimately their use of the program, the
location and size (e.g., maximum number of bicycle racks available) of
stations matters. Indeed, distance to bicycle share stations is
associated with use (Fuller et al., 2011; J. Wang and Lindsey, 2019b)
and can be a barrier to using PBSPs (Fishman et al., 2014). Kabra et
al.~(2020) found that the majority of bike share usage in Paris comes
from areas within 300 metres of stations, which amounts to 2-4 minutes
walking by an able-bodied adult. Furthermore, similar to other public
amenities affected by crowding, like health care services (e.g.,
Pereira, Carlos Kauê Vieira Braga, et al., 2021), the utility of
stations is limited by the number of bicycles that they can hold. The
program may not necessarily be improved if stations are easy to reach
but offer only a small number of bicycles. Likewise, more people may not
opt to use the program if the supply of bicycles available at the
nearest station is insufficient to meet demand. The location and size of
stations is important to increase the utility of this public
transportation option for more people, thus achieving vertical equity.
Accessibility analyses for PBSPs constitute a positive and
evaluation-based approach that also has the potential to inform equity
efforts. For instance, Wang and Lindsey (2019b) investigated whether new
or relocated bicycle share stations increased accessibility and use,
which offered important insights to improve the performance of the
program.

@@

In contrast, floating catchment area (FCA) methods incorporate
information on capacity and demand in calculating accessibility. FCA
measures have been widely employed in healthcare accessibility research
and are typically calculated across two steps. In the first, a ratio of
supply to demand at service locations is calculated, such as the number
of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centres and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. While there have been many
methodological innovations in FCA methods (for example, {\textbf{???}};
{\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}), a
recent improvement to this approach was achieved through a simple and
intuitive balancing of the impedance that addressed the effects of
demand and service inflation found in previous FCA approaches (see Paez
et al., 2019).

When measuring accessibility, researchers have taken different
approaches when it comes to aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregate data from entire
neighbourhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequities in PBSP usage.

Using the balanced floating catchment area method (Paez et al., 2019), a
novel approach that has not been used yet in cycling research, with
disaggregate population-level data, in this paper we examine
accessibility to a public bicycle share program in Hamilton, Ontario. We
(1) conduct a sensitivity analysis by measuring accessibility and level
of service to bike share stations at different walking time thresholds
to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
(2) explore the contribution of specific stations that were added to
Hamilton's PBSP to reducing both horizontal and vertical inequities; and
(3) examine whether disparities in accessibility exist according to
median household income of dissemination areas within the core service
area.

@@

In fact, since the proportional allocation procedure means that any
proportion of the population allocated to a station is never allocated
to other stations, and conversely any level of service allocated to a
population is never re-allocated elsewhere, this property is replicated
for any level of aggregation. For this paper, we employ a hybrid
location-based and person-based approach to calculating accessibility
using disaggregate population data. In their review of accessibility
measures, Geurs and van Wee (2004) highlight the need for greater
inclusion of individual spatio-temporal constraints but acknowledge the
challenges of acquiring and analyzing person-based data. This comes
after Kwan's (1998) work to show that space-time measures are better
able of capturing interpersonal differences, especially the effect of
space-time constraints on individual behaviour, and are more helpful for
unraveling gender/ethnic differences. Applying the balanced floating
catchment area approach allows us to examine accessibility by
stratifying according to median household income, which would constitute
the individual component of the accessibility measure (Geurs and van
Wee, 2004). However, conducting a further sensitivity analysis to
measure accessibility at different distance thresholds would help to
consider potential spatio-temporal constraints. Different people may be
willing to travel different distances to access a public bicycle share
station.

@@

To obtain population at sub-census geography levels (at the micro scale)
we use pycnophylactic interpolation (Tobler, 1979). We obtained
population data from the 2016 Census of Canada for dissemination areas
(DAs, the smallest publicly available census geography in Canada). This
zonal values of the population were interpolated to smaller polygons 50
by 50 metres in size. Pycnophilactic interpolation involves smoothing
out the population from each dissemination area while preserving total
volume {[}see Figure \ref{fig:da-population}{]}. When interpolating the
population at this high level of resolution it is important to ensure
that population numbers were not allocated to areas where people do not
live in Hamilton (e.g., to parks, large institutional buildings, etc.)
To do so, we retrieved shapefiles for various geographic features {[}see
Table \ref{tab:data-features}{]} from Open Hamilton. Next, we removed
these features from the PBSP core service area and used pycnophylactic
interpolation to disaggregate and reallocate population within the
remaining area {[}see Figure \ref{fig:interpolated-population}{]}.

@@

\caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). Each micro population cell is 50-by-50 m in size. The downtown area is outlined in dark green.}\label{fig:interpolated-population}
\end{figure}

@@

Consensus regarding the distance that individuals are willing to walk to
access a PBSP station is lacking, but the literature on walking
behaviour was consulted to determine the thresholds for the sensitivity
analysis. Previous studies have found that living within 250 metres
(Fuller et al., 2011) and 300 metres (Kabra et al., 2020) is correlated
with bike share use, while other research has found that walking trips
are less than 600 metres and rarely more than 1200 metres (Millward et
al., 2013) or a median distance of 650 metres (Larsen et al., 2010). In
Hamilton, Hamilton Bike Share will depict a map at some stations to show
the user the locations of the other nearest stations within a five
minute walk, which suggests that this is an average distance that people
are expected to walk.

@@

For this reason, we experiment with various walking thresholds by
conducting a sensitivity analysis to calculate accessibility at
different walking times from population cell to bicycle share station: 3
minutes, 5 minutes, 10 minutes, and 15 minutes. We categorize these
thresholds as minimum, average, maximum, and extreme, respectively. At
each threshold, we compare accessibility between the current system and
the original system to examine the contribution of the additional equity
stations.

@@

With a walking distance of three minutes, we find that there are 25.2
bicycle racks per person in the original system. The addition of equity
stations increases this ratio slightly to 25.4 bicycles per person.
Figure \ref{fig:figure-6} presents a comparison of accessibility between
the systems for the population cells. Accessibility is fairly uniform
overall, with the exception of two small areas where accessibility is
slightly higher.

@@

With a walking distance of five minutes, we find that there are 68.6
bicycle racks per person in the original system. With the addition of
equity stations, there are now 68.8 bicycles per person. Figure
\ref{fig:figure-7} presents a comparison of accessibility between the
systems. Again, accessibility is fairly uniform, with the exception of
one very small area. At this threshold, there are more bicycle racks per
person than at the minimum threshold. This occurs because fewer stations
can be reached when the travel time is 3 minutes or less.

@@

With a walking distance of ten minutes, we find that there are 3.61
bicycle racks per person in the original system. With the addition of
equity stations, there are now 3.74 bicycles per person. Figure
\ref{fig:figure-8} presents a comparison of accessibility between the
systems. We begin to see differences in accessibility across the service
area, with users near the university and its adjacent neighbourhoods, as
well as neighbourhoods north of the downtown area (outlined in green),
having slightly higher accessibility. While the differences are modest,
they are more apparent at this threshold than at shorter walking
distances.

@@

\hypertarget{accessibility-by-median-household-income}{%
\subsection{Accessibility by Median Household
Income}\label{accessibility-by-median-household-income}}

One of the unique properties of the balanced floating catchment area
method is that data can be reaggregated while preserving the total
population in the service area and the supply at each station. This
avoids demand and supply inflation, and also enables us to present
findings in a way that is easier and more intuitive to interpret.
Therefore, we reaggregate population and accessibility from the smallest
50 by 50 metres population cells to the median household income for each
dissemination area. We divide income by quintiles: bottom 20\%, second
20\%, third 20\%, fourth 20\%, and top 20\%.

Panels 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
{[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
maps with the combined spatial distribution of accessibility and median
household income at different thresholds. Our analysis demonstrates that
the extreme threshold of fifteen minutes allows the most people to
access a station. We find that stations added to Hamilton's public
bicycle share program to increase equity for disadvantaged
neighbourhoods achieved their goal by increasing access, albeit only
modestly {[}see Table \ref{tab:accessibility-income}{]}. By implementing
equity stations in more areas with lower median total income in
Hamilton, the PBSP has achieved greater horizontal equity by extending
the spatial distribution of bicycles across the city. This is
particularly evident at the minimum and average thresholds of three and
five minutes, respectively, where the equity stations fill a number of
gaps in program coverage. The largest gains were made for dissemination
areas in the second 20\%, where an additional 3,073 and 5,395 people
could reach a bicycle share station within three and five minutes walk,
respectively, after the addition of the equity station.

However, we found that dissemination areas with the lowest median
household income do not have much greater access to the program {[}see
Table \ref{tab:table-3}{]}. Income disparities still persist, however
only at certain thresholds. With and without equity stations, people in
the top 20\% of income have the highest level of access at a threshold
of ten and fifteen minutes. Although dissemination areas in the second
20\% have the highest level of access by a significant amount at lower
distance thresholds, the bottom 20\%, who may benefit the most from
Hamilton Bike Share's equity initiatives, have the lowest access at
three minutes threshold and the second lowest access at all other
thresholds.

@@

@@

\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).}

@@

\begin{tabular}[t]{lcccccc}

@@

\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Without Equity Stations} & \multicolumn{2}{c}{With Equity Stations} & \multicolumn{2}{c}{Difference} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Income Quintile & Accessibility & Population & Accessibility & Population & Accessibility & Population\\

@@

\multicolumn{7}{l}{\textbf{Threshold - 3 minutes}}\\
\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{2.377} & \cellcolor{gray!6}{22359} & \cellcolor{gray!6}{2.424} & \cellcolor{gray!6}{22798} & \cellcolor{gray!6}{0.047} & \cellcolor{gray!6}{439}\\
\hspace{1em}Second 20\% & 12.203 & 9347 & 12.281 & 12420 & 0.078 & 3073\\

>>>>

  \title{Examining spatial equity in accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) are implemented to promote cycling
  as a convenient and sustainable mode of transportation. These systems
  can also play a role in addressing transportation needs and advancing
  transportation equity if they make cycling more accessible to lower
  income or disadvantaged populations. In Ontario, the City of Hamilton
  launched a public bicycle share program in 2015 that currently has over
  900 operational bicycles and 130 docking stations. This system was
  expanded in 2018 by an equity initiative that added twelve docking
  stations with the explicit objective of increasing spatial equity in
  access. Since the cost of reaching a bicycle share station decreases the
  potential of accessing the program, the location of stations matters,
  and previous research found that Hamilton's public bicycle share program
  targeted well disadvantaged areas in the city. In this research, we
  revisit the case of Hamilton and investigate differentials in
  accessibility to bicycle share stations using a balanced floating
  catchment area (BFCA) accessibility approach. We compare accessibility
  with and without the equity stations to assess the effect of the
  initiative. We implement our analysis parting from micro zones to better
  reflect walking to a docking station, and conduct a sensitivity analysis
  at several walking time thresholds. We then reaggregate the estimated
  accessibility for further analysis using census data. Our findings
  indicate that the addition of equity stations increased the horizontal
  equity of the system by growing the serviced population at every
  threshold examined. With respect to vertical equity, although
  accessibility increased for the whole population, this increase was
  relatively modest especially for population in the bottom 20\% of median
  total household income.
  \end{abstract}

@@

The potential of public bicycle share programs (PBSPs) to increase
cycling levels is but one of many reasons for implementing such programs
in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
and convenient form of public transportation, shared bicycles can
encourage individuals to take up cycling for short local trips or first
and last mile trips to other public transportation instead of using
personal vehicles. These systems can also play a role in addressing
transportation needs and advancing transportation equity if they make
cycling more accessible to lower income or disadvantaged populations. As
these systems become increasingly common, their introduction has been
accompanied by a flurry of research that investigates them from the
perspective of spatial and transportation equity (e.g., Smith et al.,
2015; Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et
al., 2019; Qian and Jaller, 2020; Qian et al., 2020). Indeed, although
these programs are available to the general public and ought to be
accessible to any individuals who wish to use them, research on PBSPs
located in North American and European cities indicates that inequity in
spatial accessibility needs redressing.

In this paper we investigate the case of the PBSP in the City of
Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known
as Social Bicycles, or SoBi) was launched in 2015 and currently has over
900 operational bicycles and 130 docking stations. An interesting
feature of this system is that it was expanded by an equity initiative
that introduced twelve docking stations in 2018 with the explicit
objective of increasing spatial accessibility. Hamilton Bike Share was
already studied by Hosford and Winters (2018) as part of a selection of
PBSPs in Canada. These authors found that most of the cities in their
research could benefit of greater efforts to expand service to areas
with lower socio-economic populations - but Hamilton fared somewhat
better than other cities in this respect. What distinguishes our
research is the use of a balanced floating catchment area (BFCA)
accessibility approach that accounts for the supply of stations and the
potential demand from the population serviced. This requires a more
disaggregated approach than the use of census geographies (q.v., Hosford
and Winters, 2018), since many trips to a bike share station likely
happen at a level lower than even the smallest census geography. For
this reason, we implement our analysis parting from micro population
zones to better reflect the friction of walking to a docking station.
Further, we conduct a sensitivity analysis at several walking time
thresholds. In terms of spatial equity, we compare accessibility with
and without the equity stations to assess the effect of the initiative,
before reaggregating the data for further analysis using median total
household income information from the census. We find that the addition
of equity stations does increase the serviced population at every
threshold examined. However, although accessibility increased for the
whole population, this increase was relatively modest, especially for
population in the bottom 20\% of median total household income. The
analysis also identifies areas under-serviced within the core service
area of the system.

@@

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent interest in prioritizing local proximity and modes
that are suitable for local trips like walking and cycling (Levine,
2020). As a measure of the ease of reaching potential destinations
spread spatially in a given area, accessibility is relevant to PBSPs
because it can identify current inequaltiy in the provision of
infrastructure, as well as guide interventions that increase access for
groups that are under-serviced or address gaps in transportation
options. It also addresses some of the challenges of other performance
measures such as level of service within a transportation network by
measuring person-based indicators and exploring differences in use
between population groups (Páez et al., 2012).

Beyond the utility derived from using shared bikes to destinations of
value, an important aspect of accessibility associated with PBSPs is the
distance an individual must travel to reach a bicycle share station
(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
distance needed to reach a bicycle share station decreases the potential
of accessing the program and ultimately their use of the program, the
location and size (e.g., number of bicycle racks) of stations matters.
Indeed, distance to bicycle share stations is associated with use
(Fuller et al., 2011; J. Wang and Lindsey, 2019b) and can be a barrier
to using PBSPs (Fishman et al., 2014). Kabra et al.~(2020) found that
the majority of bike share usage in Paris comes from areas within 300 m
of stations, which amounts to 2-4 minutes walking by an able-bodied
adult. Furthermore, similar to other public amenities affected by
crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
Braga, et al., 2021), the utility of stations is limited by the number
of bicycles that they can hold. The program may not necessarily be
improved if stations are easy to reach but offer only a small number of
bicycles. Likewise, more people may not opt to use the program if the
supply of bicycles available at nearby stations is insufficient to meet
demand. The location and size of stations is important to increase the
utility of this public transportation option for more people, thus
achieving vertical equity. Accessibility analyses for PBSPs constitute a
positive and evaluation-based approach that also has the potential to
inform equity efforts. For instance, Wang and Lindsey (2019b)
investigated whether new or relocated bicycle share stations increased
accessibility and use, which offered important insights to improve the
performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate
information on capacity and demand in calculating accessibility. FCA
measures have been widely employed in healthcare accessibility research
and are typically calculated across two steps. In the first, a ratio of
supply to demand at service locations is calculated, such as the number
of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centres and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. While there have been many
methodological innovations in FCA methods (for example, {\textbf{???}};
{\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}), a
recent improvement to this approach was achieved through a simple and
intuitive balancing of the impedance that addressed the effects of
demand and service inflation found in earlier FCA approaches (see Paez
et al., 2019).

When measuring accessibility, researchers have taken different
approaches when it comes to aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregate data from entire
neighbourhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequity in PBSP usage.

Using the balanced floating catchment area method (Paez et al., 2019), a
novel approach that has not been used yet in cycling research, with
disaggregate population-level data, in this paper we examine
accessibility to a public bicycle share program in Hamilton, Ontario. We
(1) conduct a sensitivity analysis by measuring accessibility and level
of service to bike share stations at different walking time thresholds
to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
(2) explore the contribution of specific stations that were added to
Hamilton's PBSP to reducing both horizontal and vertical inequities; and
(3) examine whether disparities in accessibility exist according to
median total household income of dissemination areas within the core
service area.

@@

In fact, since the proportional allocation procedure means that any
proportion of the population allocated to a station is never allocated
to other stations, and conversely any level of service allocated to a
population is never re-allocated elsewhere, this property is replicated
for any level of aggregation. For this paper, we employ a hybrid
location-based and person-based approach to calculating accessibility
using disaggregate population data. In their review of accessibility
measures, Geurs and van Wee (2004) highlight the need for greater
inclusion of individual spatio-temporal constraints but acknowledge the
challenges of acquiring and analyzing person-based data. This comes
after Kwan's (1998) work to show that space-time measures are better
able of capturing interpersonal differences, especially the effect of
space-time constraints on individual behaviour, and are more helpful for
unraveling gender/ethnic differences. Applying the balanced floating
catchment area approach allows us to examine accessibility by
stratifying according to median total household income, which would
constitute the individual component of the accessibility measure (Geurs
and van Wee, 2004). However, conducting a further sensitivity analysis
to measure accessibility at different distance thresholds would help to
consider potential spatio-temporal constraints. Different people may be
willing to travel different distances to access a public bicycle share
station.

@@

To obtain population at sub-census geography levels (at the micro scale)
we use pycnophylactic interpolation (Tobler, 1979). We obtained
population data from the 2016 Census of Canada for dissemination areas
(DAs, the smallest publicly available census geography in Canada). This
zonal values of the population were interpolated to smaller polygons
50-by-50 m in size. Pycnophilactic interpolation involves smoothing out
the population from each dissemination area while preserving total
volume {[}see Figure \ref{fig:da-population}{]}. When interpolating the
population at this high level of resolution it is important to ensure
that population numbers were not allocated to areas where people do not
live in Hamilton (e.g., to parks, large institutional buildings, etc.)
To do so, we retrieved shapefiles for various geographic features {[}see
Table \ref{tab:data-features}{]} from Open Hamilton. Next, we removed
these features from the PBSP core service area and used pycnophylactic
interpolation to disaggregate and reallocate population within the
remaining area {[}see Figure \ref{fig:interpolated-population}{]}.

@@

\caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). Each population cell is 50-by-50 m in size. The downtown area is outlined in dark green.}\label{fig:interpolated-population}
\end{figure}

@@

Consensus regarding the distance that individuals are willing to walk to
access a PBSP station is lacking, but the literature on walking
behaviour can provide some guidelines to determine the thresholds for
sensitivity analysis. Previous studies have found that living within 250
metres (Fuller et al., 2011) and 300 metres (Kabra et al., 2020) is
correlated with bike share use, while other research has found that
walking trips are less than 600 metres and rarely more than 1200 metres
(Millward et al., 2013) or a median distance of 650 metres (Larsen et
al., 2010). In Hamilton, Hamilton Bike Share will depict a map at some
stations to show the user the locations of the other nearest stations
within a five minute walk, which suggests that this is an average
distance that people are expected to walk.

@@

For this reason, we experiment with various walking thresholds by
conducting a sensitivity analysis to calculate accessibility at
different walking times from population cell to bicycle share station: 3
minutes, 5 minutes, 10 minutes, and 15 minutes. We categorize these
thresholds as minimum, average, maximum, and extreme, respectively. At
each threshold, we compare accessibility between the current system and
the original system to examine the contribution of the additional equity
stations. When considering the results reported below, it is important
to remember that accessibility is technically a form of smoothing
(O'Kelly and Horner, 2003, pp. 7--8): smaller thresholds produce less
smoothing (which can result in ``spiky'' accessibility landscapes), and
larger thresholds produce more smoothing, and fewer spikes.

@@

With a walking distance of three minutes, we find that there are 25.2
bicycle racks per person system-wide in the original system
configuration (i.e., without equity stations). The addition of equity
stations increases this ratio slightly to 25.4 bicycles per person.
Figure \ref{fig:figure-6} presents a comparison of accessibility between
the systems for the population cells. Accessibility is fairly uniform
overall, with the exception of two small areas where accessibility is
slightly higher. This high level of system-wide accessibility occurs
because the population that can reach bike share stations when travel
time is 3 minutes or less is very limited, and accessibility is strongly
shaped by a few locations that concentrate population and stations.

@@

With a walking distance of five minutes, we find that there are 68.6
bicycle racks per person in the original system. With the addition of
equity stations, there are now 68.8 bicycle racks per person. At this
threshold, there are more bicycle racks per person than at the minimum
threshold. System-wide accessibility has in fact increased: the
population that can reach the stations has grown, but not to the the
point that congestion effects begin to take place. Figure
\ref{fig:figure-7} presents a comparison of accessibility between the
systems. Again, accessibility is fairly uniform, with the exception of
one very small area.

@@

With a walking distance of ten minutes, we find that there are 3.61
bicycle racks per person in the original system. With the addition of
equity stations, there are now 3.74 bicycles per person. Figure
\ref{fig:figure-8} presents a comparison of accessibility between the
systems. Differences in accessibility across the service area are now
apparent, with users near the university and its adjacent
neighbourhoods, as well as neighbourhoods north of the downtown area
(outlined in green), having slightly higher accessibility. While the
differences are modest, they are more apparent at this threshold than at
shorter walking distances.

@@

\hypertarget{accessibility-by-median-total-household-income}{%
\subsection{Accessibility by Median Total Household
Income}\label{accessibility-by-median-total-household-income}}

A unique property of the balanced floating catchment area method is that
data can be reaggregated while preserving the total population in the
service area and the supply at each station. This avoids demand and
supply inflation, and also enables us to present findings in a way that
is perhaps more intuitive to interpret. Therefore, we reaggregate
population and accessibility from the 50-by-50 m population cells to the
dissemination area. We then can compare accessibility to median total
household income statistics from the census.

Figures 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
{[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
maps with the combine the spatial distribution of accessibility and
median total household income, using tertiles for the coloring scheme.
In each figure the top panel is the system without the equity stations,
and the bottom panel is accessibility with the equity stations. As
expected, the analysis indicates that the extreme threshold of fifteen
minutes is associated with the largest number of people who are within
the assumed service areas of stations. We find that stations added to
Hamilton's public bicycle share program did indeed expand the spatial
coverage of the system. By implementing equity stations in areas with
lower median total household income in Hamilton, the PBSP attained
greater horizontal equity by extending the spatial distribution of
bicycles across the city. This is particularly evident at the minimum
and average thresholds of three and five minutes, respectively, where
the equity stations fill a number of gaps in program coverage.

@@

To further explore this issue, we aggregate accessibility by median
total household income across all geographies (see Table
\ref{tab:accessibility-income}). Similar to Hosford and Winters (2018),
we find that Hamilton's PBSP already serviced a large proportion of the
population in the bottom and second quintiles of median total household
income. This may well be an artifact of the spatial socio-economic and
demographic profile of Hamilton, where the most dense parts of the city
(where a PBSP is most easily launched) are also those with relatively
lower incomes. On the other hand, the levels of accessibility to
Hamilton's PBSP are generally lower for populations in the bottom 20\%
of median total household income, compared to populations in the top
20\%.

The results indicate that the addition of equity stations increased
horizontal equity by growing the population serviced irrespective of the
walking threshold. The largest gains were made for dissemination areas
in the second 20\%, where an additional 3,073 and 5,395 people could
reach a bicycle share station within three and five minutes walk,
respectively, after the addition of the equity station. However, we find
that there were only small changes in the population in the bottom 20\%
of median total household income who are serviced by the equity
stations, and the accessibility gains are also quite modest and smaller
than for populations in the second and third quintiles of median total
household income. This suggests that income disparities persist, albeit
this depends on the walking time thresholds. With and without equity
stations, people in the top 20\% of income have the highest level of
access at a threshold of ten and fifteen minutes. Although dissemination
areas in the second 20\% have the highest level of access by a
significant amount at lower distance thresholds, the bottom 20\%, who
may benefit the most from Hamilton Bike Share's equity initiatives, have
the lowest access at three minutes threshold and the second lowest
access at all other thresholds.

@@

\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations). Total population is the population by income quintile in the DAs that have any PBSP service at all.}

@@

\begin{tabular}[t]{lccccccl}

@@

\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Without Equity Stations} & \multicolumn{2}{c}{With Equity Stations} & \multicolumn{2}{c}{Difference} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8}
Income Quintile & Total Population & Population & Accessibility & Population & Accessibility & Population & Accessibility\\

@@

\multicolumn{8}{l}{\textbf{Threshold - 3 minutes}}\\
\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{22359} & \cellcolor{gray!6}{2.377} & \cellcolor{gray!6}{22798} & \cellcolor{gray!6}{2.424} & \cellcolor{gray!6}{439} & \cellcolor{gray!6}{0.047}\\
\hspace{1em}Second 20\% & 33312 & 9347 & 12.203 & 12420 & 12.281 & 3073 & 0.078\\
\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{7745} & \cellcolor{gray!6}{3.093} & \cellcolor{gray!6}{9455} & \cellcolor{gray!6}{3.156} & \cellcolor{gray!6}{1710} & \cellcolor{gray!6}{0.063}\\
\hspace{1em}Fourth 20\% & 20185 & 1673 & 4.119 & 1673 & 4.119 & 0 & 0.000\\
